A new tidy data structure to support exploration and modeling of temporal data

### 2019-11-21 tsibble.Rmd
typo fixes

The paper is structured as follows. Section \@ref(sec:str) reviews temporal data structures corresponding to time series and longitudinal analysis, and discusses "tidy data". Section \@ref(sec:semantics) proposes contextual semantics for temporal data, built on top of tidy data principles. The concept of data pipelines, with respect to the time domain, is discussed in depth in Section \@ref(sec:pipeline), followed by a discussion of the design choices made in the software implementation in Section \@ref(sec:software). Two case studies are presented in Section \@ref(sec:cases) illustrating temporal data exploration using the new infrastructure. Section \@ref(sec:conclusion) summarises current work and discusses future directions.

@@

Temporal data problems often fall into two types of analysis, time series and longitudinal. Both of these may have similar data input, but the representation for modelling is typically different. Time series analysis tends to focus on the dependency within series, and the cross-correlation between series. Longitudinal analysis tends to focus on overall temporal patterns across demographic or experimental treatment strata, that incorporates within subject dependency.

@@

Inspired by a "primary key" [@codd_relational_1970], a unique identifier for each observation in a relational database, the tsibble key also uniquely identifies each observational unit over time. When constructing a tsibble, any duplicates of key-index (i.e. key-value) pairs will fail, because duplicates signal a data quality issue, which would likely affect subsequent analyses and hence decision making. For example, either `gender` or `country` alone is not enough to be the key for the tuberculosis data. Analysts are encouraged to better understand the data, or reason about the process of data cleaning when handling duplicates. Figure \@ref(fig:tsibble-tidy) reveals the tidy module with clear routes required for a tsibble. The rigidity of tsibble, as the fundamental data infrastructure, ensures the validity of subsequent temporal data analysis.

@@

Since observational units are embedded, modelling and forecasting across units and time in a tsibble will be simplified. The tsibble key plays a role of the central transit hub in connecting multiple tables managed by the data, models, and forecasts. This neatly decouples the expensive data copy from downstream summaries, significantly reducing the storage space.

@@

To abide by the "tidy data" rules -- "Each type of observational units should form a table" -- in a tsibble each observational unit shares a common interval. This means that a tsibble will report one single interval, whether the data has a fixed or mixed set of intervals. To handle mixed interval data, it should be organized into separate tsibbles for a well-tailored analysis.

@@

A fluent data pipeline anticipates a standard data structure. The tsibble data abstraction lays the plumbing for data analysis modules of transformation, visualization and modeling in temporal contexts. It provides a data infrastructure to a new ecosystem, **tidyverts**. (The name "tidyverts" is a play on the term "tidyverse" that acknowledges the time series analysis purpose.)

@@

Implicit missingness are values that should be present but are absent. In regularly spaced temporal data, these are data entries that should be available at certain time points but are missing, leaving gaps in time. These can be detected when computing the interval estimate. It will be a problem for temporal models and operations like lag/lead are applied. A family of verbs are provided to help explore implicit missing values, and convert them into an explicit state, as follows:

@@

Line charts are universally accepted for ordered data, such as time series plots or spaghetti plots, depending on the fields. But they end up with exactly the same grammar: chronological time mapped to the horizontal axis, and the interested measurement on the vertical axis, for each unit. Many specialist plots centering around time series or longitudinal data, hence can be described and re-created under the umbrella of the grammar and **ggplot2**.

@@

Modeling is crucial to explanatory and predictive analytics, where time series and longitudinal data analysis diverge. The tsibble, as a model-oriented object, can flow into both types of modelling, and the new semantics (index and key) can be internally utilized to accelerate modelling.

Most time series models are univariate, such as ARIMA and Exponential Smoothing, modelling temporal dynamics for each series independently. The **fable** package [@R-fable], currently under development, provides a tidy forecasting framework built on top of tsibble, with the goal of promoting transparent and human-centered forecasting practices. With the presence of the key, a tsibble can hold many series. Since models are fundamentally scalable, the `model()` and `forecast()` generics will take care of fitting and forecasting univariate models to each series across time in a tsibble at once.

@@

Rolling windows adapt to functional programming, for which the **purrr** package [@R-purrr] sets a good example. These functions accept and return arbitrary inputs and outputs, with arbitrary methods. For example, moving averages anticipate numerics and produce averaged numerics via `mean()`. However, rolling window regression feeds a data frame into a linear regression method like `lm()`, and generates a complex object that contains coefficients, fitted values, and etc.

Rolling windows not only iterate but roll over a sequence of elements of a fixed window. A complete and consistent set of tools are available for facilitating window-related operations, a family of `slide()`, `tile()`, `stretch()`, and their variants. `slide()` expects one input, `slide2()` two inputs, and `pslide()` multiple inputs. For type stability, the functions always return lists. Other variants including `*_lgl()`, `*_int()`, `*_dbl()`, `*_chr()` return vectors of the corresponding types, as well as `*_dfr()` and `*_dfc()` for row-binding and column-binding data frames respectively. Their multiprocessing equivalents prefixed by `future_*()` enable rolling in parallel, via **future** [@R-future] and **furrr** [@R-furrr].

@@

The **tsibble** packages leverages the **tidyverse** grammars and pipelines through tidy evaluation [@tidy-eval] via the **rlang** package [@R-rlang]. In particular, the table verbs extensively use tidy evaluation to evaluate computation in the context of tsibble data and spotlights the "tidy" interface that is compatible with the **tidyverse**. This not only saves a few keystrokes without explicitly repeating references to the data source, but the resulting code is typically cleaner and more expressive, when doing interactive data analysis.

@@

The dataset of on-time performance for US domestic flights in 2017 represents event-driven data caught in the wild, sourced from US Bureau of Transportation Statistics [@flights-data]. It contains `r format(NROW(flights), big.mark = ",")` operating flights with many measurements (such as departure delay, arrival delay in minutes, and other performance metrics) and detailed flight information (such as origin, destination, plane number and etc.) in a tabular format. This kind of event describes each flight scheduled for departure at a time point in its local time zone. Every single flight should be uniquely identified by the flight number and its scheduled departure time, from a passenger's point of view. In fact, it fails to pass the tsibble hurdle due to duplicates in the original data. An error is immediately raised when attempting to convert this data into a tsibble, and a closer inspection has to be carried out to locate the issue. The **tsibble** package provides tools to easily locate the duplicates in the data with `duplicates()`. The problematic entries are shown below.

@@

Transforming a tsibble for exploratory data analysis with a suite of time-aware and general-purpose manipulation verbs can result in well-constructed pipelines. A couple of use cases are described to show how to approach the interested questions by wrangling the tsibble while maintaining its temporal context.

What time of day and day of week should passengers travel to avoid suffering from horrible delay? Figure \@ref(fig:draw-qtl) plots hourly quantile estimates across day of week in the form of small multiples. The upper-tail delay behaviors are of primary interest, and hence 50%, 80% and 95% quantiles are computed. This pipeline is initialized by regularizing and reshaping the list of the upper quantiles of departure delays for each hour. To visualize the temporal profiles, the time components (for example hours and weekdays) are extracted from the index. The flow chart (Figure \@ref(fig:flight-pipeline3)) demonstrates the operations undertaking in the data pipeline. The input to this pipeline is a tsibble of irregular interval for all flights, and the output ends up with a tsibble of one-hour interval by quantiles. To reduce the likelihood of suffering a delay, it is recommended to avoid the peak hour around 6pm (18) from Figure \@ref(fig:draw-qtl).

@@

A closer examination of some big airports across the US will give an indication of how well the busiest airports manage the outflow traffic on a daily basis. A subset that contains observations for Houston (IAH), New York (JFK), Kalaoa (KOA), Los Angeles (LAX) and Seattle (SEA) airports is obtained first. The succeeding operations compute delayed percentages every day at each airport, which are shown as grey lines in Figure \@ref(fig:sel-monthly-plot). Winter months tend to fluctuate a lot compared to the summer across all the airports. Superimposed on the plot are two-month moving averages, so the temporal trend is more visible. Since the number of days for each month is variable, moving averages over two months will require a weights input. But the weights specification can be avoided using a pair of commonly used rectangling verbs--`nest()` and `unnest()`, to wrap data frames partitioned by months into list-columns. The sliding operation with a large window size smooths out the fluctuations and gives a stable trend around 25% over the year, for IAH, JFK, LAX and SEA. LAX airport has seen a gradual decline in delays over the year, whereas the SEA airport has a steady delay. The IAH and JFK airports have more delays in the middle of year, while the KOA has the inverse pattern with higher delay percentage in both ends of the year. This pipeline gets the data into the daily series, and shifts the focus to five selected airports.

@@

Sensors have been installed in households across major cities in Australia to collect data for the smart city project. One of the trials is monitoring households' electricity usage through installed smart meters in the area of Newcastle over 2010--2014 [@smart-meter]. Data from 2013 have been sliced to examine temporal patterns of customers' energy consumption with **tsibble** for this case study. Half-hourly general supply in kwH have been recorded for `r format(tsibble::n_keys(elec_ts), big.mark = ",")` customers in the data set, resulting in `r format(NROW(elec_ts), big.mark = ",")` observations in total. Daily high and low temperatures in Newcastle in 2013 provide explanatory variables other than time in a different data table [@newcastle-temp], obtained using the R package **bomrang** [@R-bomrang]. Aggregating the half-hourly energy data to the same daily time interval as the temperature data allows us to join the two data tables to explore how local weather can contribute to the variations of daily electricity use and the accuracy of demand forecasting.

@@

The supporting toolkits articulate the temporal data pipeline, with the shared goal of reducing the time between framing of data questions and the code realization. The rapid iteration for broader understanding of the data is achieved through frictionlessly shifting among transformation, visualization, and modelling, using the standardized tsibble data infrastructure.

>>>>

The paper is structured as follows. Section \@ref(sec:str) reviews temporal data structures corresponding to time series and longitudinal analysis, and discusses "tidy data". Section \@ref(sec:semantics) proposes contextual semantics for temporal data, built on top of tidy data principles. The concept of data pipelines, with respect to the time domain, is discussed in depth in Section \@ref(sec:pipeline), followed by a discussion of the design choices made in the software implementation in Section \@ref(sec:software). Two case studies are presented in Section \@ref(sec:cases) illustrating temporal data exploration using the new infrastructure. Section \@ref(sec:conclusion) summarizes current work and discusses future directions.

@@

Temporal data problems often fall into two types of analysis, time series and longitudinal. Both of these may have similar data input, but the representation for modeling is typically different. Time series analysis tends to focus on the dependency within series, and the cross-correlation between series. Longitudinal analysis tends to focus on overall temporal patterns across demographic or experimental treatment strata, that incorporates within subject dependency.

@@

Inspired by a "primary key" [@codd_relational_1970], a unique identifier for each observation in a relational database, the tsibble key also uniquely identifies each observational unit over time. When constructing a tsibble, any duplicates of key-index pairs will fail, because duplicates signal a data quality issue, which would likely affect subsequent analyses and hence decision making. For example, either `gender` or `country` alone is not enough to be the key for the tuberculosis data. Analysts are encouraged to better understand the data, or reason about the process of data cleaning when handling duplicates. Figure \@ref(fig:tsibble-tidy) reveals the tidy module with clear routes required for a tsibble. The rigidity of tsibble, as the fundamental data infrastructure, ensures the validity of subsequent temporal data analysis.

@@

Since observational units are embedded, modeling and forecasting across units and time in a tsibble will be simplified. The tsibble key plays a role of the central transit hub in connecting multiple tables managed by the data, models, and forecasts. This neatly decouples the expensive data copy from downstream summaries, significantly reducing the storage space.

@@

To abide by the "tidy data" rules -- "Each type of observational unit should form a table" -- in a tsibble each observational unit shares a common interval. This means that a tsibble will report one single interval, whether the data has a fixed or mixed set of intervals. To handle mixed interval data, it should be organized into separate tsibbles for a well-tailored analysis.

@@

A fluent data pipeline anticipates a standard data structure. The tsibble data abstraction lays the plumbing for data analysis modules of transformation, visualization and modeling in temporal contexts. It provides a data infrastructure to a new ecosystem, **tidyverts** [@tidyverts]. (The name "tidyverts" is a play on the term "tidyverse" that acknowledges the time series analysis purpose.)

@@

Implicit missings are values that should be present but are absent. In regularly spaced temporal data, these are data entries that should be available at certain time points but are missing, leaving gaps in time. These can be detected when computing the interval estimate. It will be a problem for temporal models and operations like lag/lead, that expect consecutive time. A family of verbs is provided to help explore implicit missing values, and convert them into an explicit state, as follows:

@@

Line charts are universally accepted for ordered data, such as time series plots or spaghetti plots, depending on the fields. But they end up with exactly the same grammar: chronological time mapped to the horizontal axis, and the measurement of interest on the vertical axis, for each unit. Many specialist plots centering around time series or longitudinal data, can hence be described and re-created under the umbrella of the grammar and **ggplot2**.

@@

Modeling is crucial to explanatory and predictive analytics, where time series and longitudinal data analysis diverge. The tsibble, as a model-oriented object, can flow into both types of modeling, and the new semantics (index and key) can be internally utilized to accelerate modeling.

Most time series models are univariate, such as ARIMA and Exponential Smoothing, modeling temporal dynamics for each series independently. The **fable** package [@R-fable], currently under development, provides a tidy forecasting framework built on top of tsibble, with the goal of promoting transparent and human-centered forecasting practices. With the presence of the key, a tsibble can hold many series. Since models are fundamentally scalable, the `model()` and `forecast()` generics will take care of fitting and forecasting univariate models to each series across time in a tsibble at once.

@@

Rolling windows adapt to functional programming, for which the **purrr** package [@R-purrr] sets a good example. These functions accept and return arbitrary inputs and outputs, with arbitrary methods. For example, moving averages anticipate numerics and produce averaged numerics via `mean()`. However, rolling window regression feeds a data frame into a linear regression method like `lm()`, and generates a complex object that contains coefficients, fitted values, etc.

Rolling windows not only iterate but roll over a sequence of elements of a fixed window. A complete and consistent set of tools is available for facilitating window-related operations, a family of `slide()`, `tile()`, `stretch()`, and their variants. `slide()` expects one input, `slide2()` two inputs, and `pslide()` multiple inputs. For type stability, the functions always return lists. Other variants including `*_lgl()`, `*_int()`, `*_dbl()`, `*_chr()` return vectors of the corresponding types, as well as `*_dfr()` and `*_dfc()` for row-binding and column-binding data frames respectively. Their multiprocessing equivalents prefixed by `future_*()` enable rolling in parallel, via **future** [@R-future] and **furrr** [@R-furrr].

@@

The **tsibble** packages leverages the **tidyverse** grammars and pipelines through tidy evaluation [@tidy-eval] via the **rlang** package [@R-rlang]. In particular, the table verbs extensively use tidy evaluation to evaluate computation in the context of tsibble data and spotlight the "tidy" interface that is compatible with the **tidyverse**. This not only saves a few keystrokes without explicitly repeating references to the data source, but the resulting code is typically cleaner and more expressive, when doing interactive data analysis.

@@

The dataset of on-time performance for US domestic flights in 2017 represents event-driven data caught in the wild, sourced from US Bureau of Transportation Statistics [@flights-data]. It contains `r format(NROW(flights), big.mark = ",")` operating flights with many measurements (such as departure delay, arrival delay in minutes, and other performance metrics) and detailed flight information (such as origin, destination, plane number, etc.) in a tabular format. This kind of event describes each flight scheduled for departure at a time point in its local time zone. Every single flight should be uniquely identified by the flight number and its scheduled departure time, from a passenger's point of view. In fact, it fails to pass the tsibble hurdle due to duplicates in the original data. An error is immediately raised when attempting to convert this data into a tsibble, and a closer inspection has to be carried out to locate the issue. The **tsibble** package provides tools to easily locate the duplicates in the data with `duplicates()`. The problematic entries are shown below.

@@

Transforming a tsibble for exploratory data analysis with a suite of time-aware and general-purpose manipulation verbs can result in well-constructed pipelines. A couple of use cases are described to show how to approach the questions of interest by wrangling the tsibble while maintaining its temporal context.

What time of day and day of week should passengers travel to avoid suffering from horrible delay? Figure \@ref(fig:draw-qtl) plots hourly quantile estimates across day of week in the form of small multiples. The upper-tail delay behaviors are of primary interest, and hence 50%, 80% and 95% quantiles are computed. This pipeline is initialized by regularizing and reshaping the list of the upper quantiles of departure delays for each hour. To visualize the temporal profiles, the time components (for example hours and weekdays) are extracted from the index. The flow chart (Figure \@ref(fig:flight-pipeline3)) demonstrates the operations undertaken in the data pipeline. The input to this pipeline is a tsibble of irregular interval for all flights, and the output ends up with a tsibble of one-hour interval by quantiles. To reduce the likelihood of suffering a delay, it is recommended to avoid the peak hour around 6pm (18) from Figure \@ref(fig:draw-qtl).

@@

A closer examination of some big airports across the US will give an indication of how well the busiest airports manage the outflow traffic on a daily basis. A subset that contains observations for Houston (IAH), New York (JFK), Kalaoa (KOA), Los Angeles (LAX) and Seattle (SEA) airports is obtained first. The succeeding operations compute delayed percentages every day at each airport, which are shown as gray lines in Figure \@ref(fig:sel-monthly-plot). Winter months tend to fluctuate a lot compared to the summer across all the airports. Superimposed on the plot are two-month moving averages, so the temporal trend is more visible. Since the number of days for each month is variable, moving averages over two months will require a weights input. But the weights specification can be avoided using a pair of commonly used rectangling verbs--`nest()` and `unnest()`, to wrap data frames partitioned by months into list-columns. The sliding operation with a large window size smooths out the fluctuations and gives a stable trend around 25% over the year, for IAH, JFK, LAX and SEA. LAX airport has seen a gradual decline in delays over the year, whereas the SEA airport has a steady delay. The IAH and JFK airports have more delays in the middle of year, while the KOA has the inverse pattern with higher delay percentage at both ends of the year. This pipeline gets the data into the daily series, and shifts the focus to five selected airports.

@@

Sensors have been installed in households across major cities in Australia to collect data for the smart city project. One of the trials is monitoring households' electricity usage through installed smart meters in the area of Newcastle over 2010--2014 [@smart-meter]. Data from 2013 have been sliced to examine temporal patterns of customers' energy consumption with **tsibble** for this case study. Half-hourly general supply in kwH has been recorded for `r format(tsibble::n_keys(elec_ts), big.mark = ",")` customers in the data set, resulting in `r format(NROW(elec_ts), big.mark = ",")` observations in total. Daily high and low temperatures in Newcastle in 2013 provide explanatory variables other than time in a different data table [@newcastle-temp], obtained using the R package **bomrang** [@R-bomrang]. Aggregating the half-hourly energy data to the same daily time interval as the temperature data allows us to join the two data tables to explore how local weather can contribute to the variations of daily electricity use and the accuracy of demand forecasting.

@@

The supporting toolkits articulate the temporal data pipeline, with the shared goal of reducing the time between framing of data questions and the code realization. The rapid iteration for broader understanding of the data is achieved through frictionlessly shifting among transformation, visualization, and modeling, using the standardized tsibble data infrastructure.
